---
title: "Coursera_Assignment 1"
author: "Sahil"
---

## COURSERA
## REPRODUCIBLE RESEARCH
#### Week 2: Assignment 1

#### Load library
```{r}
library(ggplot2)
```

#### Loading data
Load data using the **read.csv()** and store it into an object called **activity**. *It is assumed that the dataset has been downloaded and saved in the current working directory*. 
```{r}
activity <- read.csv("C:\\Users\\Sahil\\OneDrive\\Documents\\Coursera\\Data Science Specialization_John Hopkins\\05_Reproducible Research\\Week 2\\activity.csv")
```

#### Data Exploration and Data Preprocessing
Check the structure, number of rows/columns and display the first 3 rows.
```{r}
str(activity)
head(activity,3)
```
There are 3 variables with 17568 observations. 


We convert **date** variable to *date class* from *factor class* for our analysis. 
```{r}
activity$date <- as.Date(activity$date)
```

Check if there are any missing values. 
```{r}
colSums(is.na(activity))
```
There are 2304 missing values in the dataset. 


### Question 1: What is the mean total number of steps taken per day?
For the first question; we will not treat the missing values.
Calculate the total steps taken per day and store it in a new data frame called **totalSteps**. 
```{r}
totalSteps <- aggregate(steps~date,activity,sum)
```
Let us check the **totalSteps** data frame. 
```{r}
head(totalSteps)
```
Create histogram to display the total steps taken per day
```{r}
plot1 <- ggplot(totalSteps, aes(date,steps)) + geom_bar(stat="identity",fill="royal blue") + xlab("Date") + ylab("Total Steps") + ggtitle("Total Number of Steps Taken Per Day") + theme(plot.title=element_text(size=14, face="bold"),axis.text=element_text(size=14), axis.title=element_text(size=14,face="bold")) + theme_light()
```
Display the histogram
```{r}
plot1
```

Calculate the mean and median for total number of steps taken per day

* Mean
```{r}
mean(totalSteps$steps)
```

* Median
```{r}
median(totalSteps$steps)
```


### Question 2: What is the average daily activity pattern?
Calculate the average number of steps taken by intervals of 5-minute intervals and store it in a new data frame called **timeSeries**.
```{r}
timeSeries <- aggregate(steps~interval,activity,mean)
```
Let us check the **timeSeries** data frame
```{r}
head(timeSeries)
```
Generate plot to display the average number of steps taken by 5 minute intervals
```{r}
plot2 <- ggplot(timeSeries,aes(interval,steps)) + geom_line(color="royal blue",size=1) + xlab("5 Minute Interval") + ylab("Steps") + ggtitle("Time Series Plot of 5 Minute Interval") + theme(plot.title=element_text(size=14, face="bold"),axis.text=element_text(size=14), axis.title=element_text(size=14,face="bold")) + theme_light()
```
Display plot.
```{r}
plot2
```

Calculate which 5 minute interval on average across all days has the maximum number of steps
```{r}
maxInterval <- timeSeries[timeSeries$steps == max(timeSeries$steps),]
maxInterval
```
The 5 minute interval on average across all days that has the maximum number of steps is 835. 

### Question 3: Impute missing values
1. Calculate the number of missing values in the dataset.
```{r}
colSums(is.na(activity))
```
The total number of missing values are 2304. They are coded as *NA*. 

2. Devise a strategy to replace the missing values in the dataset. 
We will replace the missing values by the mean. 

3. Create a new dataset that is equal to the original dataset with missing data filled in. 

The new dataset is called **newActivity**.
```{r}
newActivity <- activity
```
Replace the missing values with the mean. 
```{r}
newActivity$steps[is.na(newActivity$steps)] <- mean(newActivity$steps,na.rm=TRUE)
```
Check the first 6 rows of **newActivity**.
```{r}
head(newActivity)
```
Missing values have been replaced by the mean. 

4. Generate a histogram of the total number of steps taken per day
Calculate the total steps taken per day and store it in a new data frame called **totalStepsImpute**. 
```{r}
totalStepsImpute <- aggregate(steps~date,newActivity,sum)
```
Let us check the **totalStepsImpute** data frame
```{r}
head(totalStepsImpute)
```
Generate histogram
```{r}
plot3 <- ggplot(totalStepsImpute, aes(date,steps)) + geom_bar(stat="identity",fill="royal blue") + xlab("Date") + ylab("Total Steps") + ggtitle("Total Number of Steps Taken Per Day") + theme(plot.title=element_text(size=14, face="bold"),axis.text=element_text(size=14), axis.title=element_text(size=14,face="bold")) + theme_light()
```
Display histogram:
```{r}
plot3
```

Calculate the mean and median for total number of steps taken per day

* Mean
```{r}
mean(totalStepsImpute$steps)
```

* Median
```{r}
median(totalStepsImpute$steps)
```
**Observations**

* Do these values differ from the estimates from the first part of the assignment?

  Measure | With Missing Values | Without Missing Values
  --------|---------------------|-----------------------
  mean    | 10766.19            | 10766.19
  median  | 10765               | 10766.19

Mean values do not differ but the median value has a slight difference.

* What is the impact of the missing data on the estimates of the total daily number of steps?

Comparing *plot1* and *plot2*; the shape of the histogram remains same. Impact of imputing missing data on the estimates is low. 

### Question 4: Are there any differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Create a new variable called **day**. It displays whether the day from the date is a *weekend* or a *weekday*. 
```{r}
newActivity$day <- as.factor(ifelse(weekdays(newActivity$date) %in% c("Saturday","Sunday"),"Weekend","Weekday"))
```
Check the first 6 rows of **newActivity**
```{r}
head(newActivity)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
Calculate the average number of steps taken across all *day* type variable
```{r}
timeSeriesDay <- aggregate(steps~interval+day,newActivity,mean)
```
Display the **timeSeriesDay** data frame
```{r}
head(timeSeriesDay)
```
Generate time series plot. 
```{r}
plot4 <- ggplot(timeSeriesDay,aes(interval,steps)) + geom_line(color="royal blue",size=1) + facet_grid(day~.) + xlab("5 Minute Interval") + ylab("Average Number of Steps Taken") + ggtitle("Average Number of Steps Taken Across Weekday/Weekend") + theme(plot.title=element_text(size=14, face="bold"),axis.text=element_text(size=14), axis.title=element_text(size=14,face="bold")) + theme_light()
```
Display plot. 
```{r}
plot4
```

**Observations** 

* Are there any differences in activity patterns between weekdays and weekends?

Plot4 depicts differences in activity patterns between weekdays and weekends. Weekday has the highest peak but weekend has more peaks compared to weekdays. Activity seems more during the weekends. 
